Ecosystem size and complexity are extrinsic drivers of food chain length in branching networks

### 2022-03-31 rmd/manuscript.Rmd
edit: manuscript

The origin of variation in food chain length (FCL), the number of feeding links from the basal species to top predators, has intrigued ecologists for decades due to its importance in regulating top-down effects [@paceTrophicCascadesRevealed1999], energy flow [@wangBiodiversityEcosystemFunctioning2018], and contaminant concentrations in top predators that humans often consume [@kiddEffectsTrophicPosition1998]. Hypotheses regarding controls of food chain length can be organized into three broad categories. First, the resource availability hypothesis states that basal productivity controls food chain length because imperfect energy conversion through predator-prey interactions restricts energy available to higher trophic levels [@oksanenExploitationEcosystemsGradients1981]. Second, the disturbance hypothesis (or the dynamical stability hypothesis) predicts longer food chains are vulnerable to environmental perturbations (i.e., the dynamical constraint). Thus, habitats with frequent and/or severe disturbance should support fewer trophic levels [@pimmNumberTrophicLevels1977]. Lastly, the ecosystem size hypothesis posits that larger ecosystems should harbor longer food chains [@takimotoEffectsProductivityDisturbance2012; @teruiSpatialDisturbanceSynchrony2019; @postEcosystemSizeDetermines2000].

Among the three hypotheses, the ecosystem size hypothesis has been consistently supported in empirical explorations, likely because multiple mechanisms may underlie the positive association [@takimotoEnvironmentalDeterminantsFoodchain2013]. For example, large ecosystems have increased species richness (species-area relationship [@lomolinoEcologyMostGeneral2000]), buffer against disturbances [@saboRoleDischargeVariation2010], and have greater numbers of immigrant sources in the ecosystem with stable metapopulations [@takimotoEnvironmentalDeterminantsFoodchain2013]. Meanwhile, the theoretical basis for the other two hypotheses rest on implicit assumptions of linear food chains, which are often violated in natural systems. As such, the effects of disturbance and resource availability on food chain length have been found to be highly variable among ecosystem types and geographic regions [@takimotoEffectsProductivityDisturbance2012; @diehlIntraguildPreySuffer2001; @takimotoEnvironmentalDeterminantsFoodchain2013; @postLongShortFoodchain2002].

More recently, food web ecologists have begun to recognize that FCL rarely responds to a single driver, but instead is determined by interactions of multiple factors. For example, @wardMechanisticTheoryAquatic2017 showed that FCL was more responsive to increasing productivity in large ecosystems compared with small ecosystems, whereas FCL was more sensitive to size in low productivity systems. Such context dependent responses to extrinsic factors seem to be tightly coupled with the arrangement of trophic links within a food web. For example, omnivory (feeding at multiple trophic levels) is a strong intrinsic driver of food chain length [@takimotoEffectsProductivityDisturbance2012; @postProximateStructuralMechanisms2007; @woottonOmnivoryStabilityFreshwater2017]. When omnivory is absent in three-species communities, the trophic structure assumes a simple linear food chain with maximum length equal to 3 (**Figure 1D**). However, as omnivory increases, FCL decreases proportionally as the top predator eats lower in the food web [@wardMechanisticTheoryAquatic2017; @woottonOmnivoryStabilityFreshwater2017]. Theoretical work predicts that food webs are destabilized when omnivory is prevalent, leading to the loss of species and shorter food chains even in productive, large ecosystems. Conversely, weak omnivory stabilize the food web and increase persistence of species within a community, particularly at intermediate levels of productivity and high disturbance levels [@woottonOmnivoryStabilityFreshwater2017; @woottonManyWeakInteractions2016]. Due to the context-dependency of FCL to multiple drivers, and the potential for interactive effects, a unified framework of multiple food chain drivers is needed to better understand when and where FCL will respond to these extrinsic and intrinsic factors.

While recent research efforts of food web ecologists have substantially advanced our understanding of food chain drivers, most of the previous work has been conducted in simple, two-dimensional systems, such as lakes [@postEcosystemSizeDetermines2000; @doiResourceAvailabilityEcosystem2009] and oceanic islands [@takimotoEcosystemSizeNot2008; @pringlePredatorinducedCollapseNiche2019]. However, many natural habitats have a high degree of spatial complexity that cannot be represented simply by ecosystem size. Branching ecosystems, for example, are ubiquitous yet overlooked landscape structures (e.g., rivers, trees, caves, mountain ranges) [@grantLivingBranchesPopulation2007; @altermattDiversityRiverineMetacommunities2013]. While we know that food web interactions are spatially influenced [@calcagnoConstraintsFoodChain2011; @pillaiMetacommunityTheoryExplains2011; @takimotoEffectsProductivityDisturbance2012; @teruiSpatialDisturbanceSynchrony2019], the theory to account for it has lagged, particularly in branching habitat architectures. Here, we theoretically explore drivers of food chain length in rivers, an excellent example of branching ecosystems.

Ample evidence suggests that branching structure controls environmental heterogeneity and dispersal pathways of organisms within river ecosystems [@altermattDiversityRiverineMetacommunities2013; @tonkinRoleDispersalRiver2018]. Water and materials propagate downstream as small streams hierarchically join to form larger streams and rivers. Consequently, environmental signals are highly variable among tributaries while showing a strong within-tributary correlation e.g., flood disturbance [@nakamuraDisturbanceRegimesStream2000; @isaakSlowClimateVelocities2016; @mcguireNetworkAnalysisReveals2014; @abbottUnexpectedSpatialStability2018]. Confluences serve as merging points where tributary dynamics aggregate to form distinct environmental conditions in downstream habitat patches. Habitat heterogeneity therefore increases with branching complexity (**Figure 1A**), which we define here as the probability of branching per unit river distance. Hence, highly-branched river networks provide diverse habitats that buffer the impact of large-scale environmental fluctuations, broadening opportunities for species to recolonize habitat patches post-disturbance [@teruiMetapopulationStabilityBranching2018]. Thus, ecosystem complexity in river networks may be a widespread yet overlooked extrinsic environmental driver which needs to be incorporated to understand food chain length.

We develop a model in which trophic community dynamics within branching habitats are simulated. We investigate extrinsic controls such as ecosystem size (number of patches), complexity (branching probability), and disturbance regime. In addition, our simulation model incorporates intrinsic drivers of food chain length including food web structure, productivity level, and dispersal ability, allowing us to investigate the context dependency on drivers of food chain length across extrinsic factors. We find that ecosystem size and complexity are both positively related to FCL, but the form of this relationship is dependent on other extrinsic and intrinsic factors. Our model is an important step in unifying drivers of FCL across extrinsic and intrinsic drivers and parsing their context-dependency in spatially complex ecosystems.

@@

***Food chain length.*** At each patch and time step, food chain length was determined as the highest trophic position ($TP_{x}(t)$) of the existing species. When the intraguild predator exists, the trophic position was determined as follows:

@@

Under each scenario, we ran `r max(sim_main_result$n_rep)` simulation replicates with `r n_sim` timesteps of trophic dynamics (including `r n_sim - unique(sim_main_result$n_timestep)` steps of warmup and burn-in). During the warmup period (`r unique(sim_main_result$n_warmup)` time steps), we introduced propagules into habitat patches every 10 time steps to allow species to establish populations. Propagules are random draws from $Poisson(0.5)$. Subsequently, we ran a burn-in period with no propagules for `r unique(sim_main_result$n_burnin)` time steps to remove the effects of initial conditions. We obtained the mean food chain length across all habitat patches for the last `r unique(sim_main_result$n_timestep)` time steps. Note that simulation replicates are differ in ecosystem size ($N_{p}\sim Unif(10,150)$) and ecosystem complexity ($P_{b} \sim Unif(0.01, 0.99)$) so we can estimate their effects on food chain length.

@@

Our simulation highlighted the interactive nature of putative controls of food chain length. In particular, disturbance regimes had the strongest influence on the qualitative relationships between FCL and ecosystem properties. Below, we report our results for low and high disturbance regimens separately.

@@

In ecosystems with a low disturbance regime, FCL was largely determined by omnivory with limited effects of ecosystem size and complexity (colored lines in **Figure 2**). Unsurprisingly, FCL was negatively related to the degree of omnivory, with no-omnivory food chains ($P$ only consumes $C$) having the longest FCL, and strong omnivory ($P$ consumes very little $C$) having the shortest FCL. These trends were consistent across dispersal scenarios (**Figure 2**).

Although the effects of ecosystem properties on FCL were minimal, there were noticeable effects on the occupancies of constituent species in some scenarios. For example, the occupancies of $C$ and $P$ increased with ecosystem size when omnivory was strong (see the most left column in **Figures S1 and S2**). However, these changes were not translated into changes in FCL due to the structural constraint imposed by omnivory ($C$ and $P$ occupy a similar trophic position; **Figure 1D**). Similarly, the occupancy of $P$ decreased with ecosystem complexity when omnivory was strong, but FCL remained almost the same as the increase of $C$ could compensate for the loss of $P$ for the same reason (**Figure S3 and S4**).

@@

A high disturbance regime produced much different responses of FCL. FCL increased across both gradients of ecosystem size and complexity under all combinations of productivity, the degree of omnivory, and dispersal distance (**Figure 3**). However, the effects of ecosystem size and complexity were mediated by ecological factors differently. Ecosystem size effects were stronger with longer dispersal distance ($\theta = 0.1$), while the opposite was true for ecosystem complexity (**Figure 3**). In addition, the effect of ecosystem complexity was pronounced in food webs with weaker omnivory (**Figure 3B**).

Interestingly, the positive effects of ecosystem size and complexity were driven by different mechanisms. For ecosystem size, the underlying mechanism was consistent when productivity was high; increased occupancies of all the three species resulted in longer food chains. However, when productivity was low, the mechanisms were more complex, particularly with strong omnivory (the third column in **Figure S1 and S2**). In this scenario, the occupancies of $B$ and $C$ decreased at larger ecosystem sizes, while $P$ increased its occupancy (**Figure S1 and S2**). Therefore, $P$ compensated for the loss of $B$ and $C$ to maintain the positive relationship between FCL and ecosystem size. For ecosystem complexity, all the three species increased with ecosystem complexity when productivity was high, translating into longer food chains in more complex ecosystems (**Figure S3 and S4**). However, when food webs contained strong omnivory with low productivity (the third column in **Figure S3 and S4**), the increased occupacy of $C$ maintained the positive relationship between FCL and ecosystem complexity by offsetting the loss of $P$. Hence, the contrasting mechanisms underlie the positive effects of ecosystem size and complexity on FCL.

@@

Discussion to be added

>>>>

The origin of variation in food chain length (FCL), the number of feeding links from the basal species to top predators, has intrigued ecologists for decades due to its importance in regulating top-down effects [@paceTrophicCascadesRevealed1999], energy flow [@wangBiodiversityEcosystemFunctioning2018], and contaminant concentrations in top predators that humans often consume [@kiddEffectsTrophicPosition1998]. Hypotheses regarding controls of FCL can be organized into three broad categories. First, the resource availability hypothesis states that basal productivity controls FCL because imperfect energy conversion through predator-prey interactions restricts energy available to higher trophic levels [@oksanenExploitationEcosystemsGradients1981]. Second, the disturbance hypothesis (or the dynamical stability hypothesis) predicts longer food chains are vulnerable to environmental perturbations (i.e., the dynamical constraint). Thus, habitats with frequent and/or severe disturbance should support fewer trophic levels [@pimmNumberTrophicLevels1977]. Lastly, the ecosystem size hypothesis posits that larger ecosystems should harbor longer food chains [@takimotoEffectsProductivityDisturbance2012a; @teruiSpatialDisturbanceSynchrony2019; @postEcosystemSizeDetermines2000].

Among the three hypotheses, the ecosystem size hypothesis has been consistently supported in empirical explorations, likely because multiple mechanisms may underlie the positive association [@takimotoEnvironmentalDeterminantsFoodchain2013]. For example, large ecosystems have increased species richness (species-area relationship [@lomolinoEcologyMostGeneral2000]), buffer against disturbances [@saboRoleDischargeVariation2010], and have greater numbers of immigrant sources in the ecosystem with stable metapopulations [@takimotoEnvironmentalDeterminantsFoodchain2013]. Meanwhile, the theoretical basis for the other two hypotheses rest on implicit assumptions of linear food chains, which are often violated in natural systems. As such, the effects of disturbance and resource availability on FCL have been found to be highly variable among ecosystem types and geographic regions [@takimotoEffectsProductivityDisturbance2012a; @diehlIntraguildPreySuffer2001; @takimotoEnvironmentalDeterminantsFoodchain2013; @postLongShortFoodchain2002].

More recently, food web ecologists have begun to recognize that FCL rarely responds to a single driver, but instead is determined by interactions of multiple factors. For example, @wardMechanisticTheoryAquatic2017 showed that FCL was more responsive to increasing productivity in large ecosystems compared with small ecosystems, whereas FCL was more sensitive to size in low productivity systems. Such context dependent responses to extrinsic factors seem to be tightly coupled with the arrangement of trophic links within a food web. For example, omnivory (feeding at multiple trophic levels) is a strong intrinsic driver of FCL [@takimotoEffectsProductivityDisturbance2012a; @postProximateStructuralMechanisms2007; @woottonOmnivoryStabilityFreshwater2017]. When omnivory is absent in three-species communities, the trophic structure assumes a simple linear food chain with maximum length equal to 3 (**Figure 1D**). However, as omnivory increases, FCL decreases proportionally as the top predator eats lower in the food web [@wardMechanisticTheoryAquatic2017; @woottonOmnivoryStabilityFreshwater2017]. Theoretical work predicts that food webs are destabilized when omnivory is prevalent, leading to the loss of species and shorter food chains even in productive, large ecosystems. Conversely, weak omnivory stabilize the food web and increase persistence of species within a community, particularly at intermediate levels of productivity and high disturbance levels [@woottonOmnivoryStabilityFreshwater2017; @woottonManyWeakInteractions2016]. Due to the context-dependency of FCL to multiple drivers, and the potential for interactive effects, a unified framework of multiple food chain drivers is needed to better understand when and where FCL will respond to these extrinsic and intrinsic factors.

While recent research efforts of food web ecologists have substantially advanced our understanding of food chain drivers, most of the previous work has been conducted in simple, two-dimensional systems, such as lakes [@postEcosystemSizeDetermines2000; @doiResourceAvailabilityEcosystem2009] and oceanic islands [@takimotoEcosystemSizeNot2008; @pringlePredatorinducedCollapseNiche2019]. However, many natural habitats have a high degree of spatial complexity that cannot be represented simply by ecosystem size. Branching ecosystems, for example, are ubiquitous yet overlooked landscape structures (e.g., rivers, trees, caves, mountain ranges) [@grantLivingBranchesPopulation2007; @altermattDiversityRiverineMetacommunities2013]. While we know that food web interactions are spatially influenced [@calcagnoConstraintsFoodChain2011; @pillaiMetacommunityTheoryExplains2011; @takimotoEffectsProductivityDisturbance2012a; @teruiSpatialDisturbanceSynchrony2019], the theory to account for it has lagged, particularly in branching habitat architectures. Here, we theoretically explore drivers of FCL in rivers, an excellent example of branching ecosystems.

Ample evidence suggests that branching structure controls environmental heterogeneity and dispersal pathways of organisms within river ecosystems [@altermattDiversityRiverineMetacommunities2013; @tonkinRoleDispersalRiver2018]. Water and materials propagate downstream as small streams hierarchically join to form larger streams and rivers. Consequently, environmental signals are highly variable among tributaries while showing a strong within-tributary correlation e.g., flood disturbance [@nakamuraDisturbanceRegimesStream2000; @isaakSlowClimateVelocities2016; @mcguireNetworkAnalysisReveals2014; @abbottUnexpectedSpatialStability2018]. Confluences serve as merging points where tributary dynamics aggregate to form distinct environmental conditions in downstream habitat patches. Habitat heterogeneity therefore increases with branching complexity (**Figure 1A**), which we define here as the probability of branching per unit river distance. Hence, highly-branched river networks provide diverse habitats that buffer the impact of large-scale environmental fluctuations, broadening opportunities for species to recolonize habitat patches post-disturbance [@teruiMetapopulationStabilityBranching2018]. Thus, ecosystem complexity in river networks may be a widespread yet overlooked extrinsic environmental driver which needs to be incorporated to understand FCL.

We develop a model in which trophic community dynamics within branching habitats are simulated. We investigate extrinsic controls such as ecosystem size (number of patches), complexity (branching probability), and disturbance regime. In addition, our simulation model incorporates intrinsic drivers of FCL including food web structure, productivity level, and dispersal ability, allowing us to investigate the context dependency on drivers of FCL across extrinsic factors. We find that ecosystem size and complexity are both positively related to FCL, but the form of this relationship is dependent on other extrinsic and intrinsic factors. Our model is an important step in unifying drivers of FCL across extrinsic and intrinsic drivers and parsing their context-dependency in spatially complex ecosystems.

@@

***Food chain length.*** At each patch and time step, FCL was determined as the highest trophic position ($TP_{x}(t)$) of the existing species. When the intraguild predator exists, the trophic position was determined as follows:

@@

Under each scenario, we ran `r max(sim_main_result$n_rep)` simulation replicates with `r n_sim` timesteps of trophic dynamics (including `r n_sim - unique(sim_main_result$n_timestep)` steps of warmup and burn-in). During the warmup period (`r unique(sim_main_result$n_warmup)` time steps), we introduced propagules into habitat patches every 10 time steps to allow species to establish populations. Propagules are random draws from $Poisson(0.5)$. Subsequently, we ran a burn-in period with no propagules for `r unique(sim_main_result$n_burnin)` time steps to remove the effects of initial conditions. We obtained the mean FCL across all habitat patches for the last `r unique(sim_main_result$n_timestep)` time steps. Note that simulation replicates are differ in ecosystem size ($N_{p}\sim Unif(10,150)$) and ecosystem complexity ($P_{b} \sim Unif(0.01, 0.99)$) so we can estimate their effects on FCL.

@@

Our simulation highlighted the interactive nature of putative controls of FCL. In particular, disturbance regimes had the strongest influence on the qualitative relationships between FCL and ecosystem properties. Below, we report our results for low and high disturbance regimens separately.

@@

In ecosystems with a low disturbance regime, FCL was largely determined by omnivory with limited effects of ecosystem size and complexity (colored lines in **Figure 2**). FCL was negatively related to the degree of omnivory, with no-omnivory food chains ($P$ only consumes $C$) having the longest FCL, and strong omnivory ($P$ consumes very little $C$) having the shortest FCL. The primary mechanism for the decreased FCL was different in low and high productivity scenarios. With low productivity, FCL became shorter with stronger omnivory because it intensifies competition between consumer and predator species. This effect can be seen in the decreased occupancies of consumer and predator species in a low productivity scenario (most left column in **Figure S1-S4**). However, when productivity was high, the occupancies of both species were \~1.0, indicating that the structural constraint of strong omnivory made FCL shorter (second left column in **Figure S1-S4**). Productivity played a secondary role in increasing FCL, particularly when omnivory was strong (compare left and right panels in **Figure 2**). Dispersal had no apparent effects on FCL (compare top and bottom panels in **Figure 2**). These trends were consistent across dispersal scenarios (**Figure 2**).

Although the effects of ecosystem properties on FCL were minimal, there were noticeable effects on the occupancies of constituent species in some scenarios. For example, the occupancies of $C$ and $P$ increased with ecosystem size when omnivory was strong (see the most left column in **Figures S1 and S2**). However, these changes were not translated into FCL because $C$ and $P$ occupy a similar trophic position (**Figure 1D**). Similarly, the occupancy of $P$ decreased with ecosystem complexity when omnivory was strong, but FCL remained almost the same as the increase of $C$ could compensate for the loss of $P$ for the same reason (**Figure S3 and S4**).

@@

On average, increased disturbance frequency decreased FCL, and omnivory had similar effects on FCL (**Figures 3**). More importantly, a high disturbance regime produced much different responses of FCL to ecosystem properties. FCL increased across both gradients of ecosystem size and complexity under all combinations of productivity, omnivory, and dispersal distance (**Figure 3**). The effect of ecosystem complexity was generally clearer than that of ecosystem size. It is important to note that the positive relationships emerged in a high disturbance regime because smaller and simpler ecosystems were disproportionately vulnerable to increased disturbance frequency (compare far left in **Figures 2 and 3**).

Dispersal buffered the disturbance effects on FCL, and the magnitude of buffering differed by ecosystem structure. The greater buffering effect of dispersal emerged in larger ecosystems (compare top and bottom panels in **Figure 3**) because greater dispersal capability is required to reach distant patches. In the meantime, simpler, less branched ecosystems benefited more from increased dispersal because habitat patches are more isolated from one another in linearly arranged networks (**Figure 3**). Productivity increased FCL as observed in a low disturbance regime, but it did not modulate the qualitative relationships between FCL and ecosystem properties (compare left and right panels in **Figure 3**).

The positive effects of ecosystem size and complexity were driven by different processes. For ecosystem size, the underlying process was consistent when productivity was high; increased occupancies of all the three species resulted in longer food chains. However, when productivity was low, the processes were more complex, particularly with strong omnivory (the third column in **Figure S1 and S2**). In this scenario, the occupancies of $B$ and $C$ decreased at larger ecosystem sizes, while $P$ increased its occupancy (**Figure S1 and S2**). Therefore, $P$ compensated for the loss of $B$ and $C$ to maintain the positive relationship between FCL and ecosystem size. For ecosystem complexity, all the three species increased with ecosystem complexity when productivity was high, translating into longer food chains in more complex ecosystems (**Figure S3 and S4**). However, when food webs contained strong omnivory with low productivity (the third column in **Figure S3 and S4**), the increased occupancy of $C$ maintained the positive relationship between FCL and ecosystem complexity by offsetting the loss of $P$. Hence, the contrasting mechanisms underlie the positive effects of ecosystem size and complexity on FCL.

@@

The ecosystem size hypothesis of FCL has received support from both theoretical and empirical research [@vanderzandenPatternsFoodChain1999; @postEcosystemSizeDetermines2000; @takimotoEffectsProductivityDisturbance2012a], although empirical work so far has rarely elaborated the underlying drivers [but see @saboRoleDischargeVariation2010; @wardMechanisticTheoryAquatic2017]. In contrast, ecosystem complexity has not been incorporated in theoretical explorations of FCL to date. Here, we show that FCL responds positively to both increasing ecosystem size and complexity. This supports our hypothesis that ecosystem complexity may represent an overlooked yet important extrinsic control of FCL. However, the relationship between FCL and ecosystem complexity varied based on additional extrinsic and intrinsic factors, supporting the emerging paradigm that there is no one single driver of FCL in ecosystems [@mchughDualInfluencesEcosystem2010; @wardMechanisticTheoryAquatic2017]. Although we focus on river networks here, we expect our results to be transferable to other branching structures, such as mountain ranges and cave systems.

While increased disturbance frequency consistently reduced FCL as predicted by dynamical stability hypothesis [@pimmNumberTrophicLevels1977], more important findings in this study were evident from their synergistic effects with ecosystem structure. Our low disturbance regime scenario can be viewed as habitat networks having nearly homogeneous environments, in which long food chains were generally supported independent of ecosystem size and complexity. Consequently, FCL was predominantly determined by food web structure. However, under high disturbance regimes, only large and complex networks could provide diverse and heterogeneous habitats with potential refugia, allowing species to thrive and maintain long food chains. While empirical studies rarely looked at the influence of ecosystem structure on FCL, several studies provide some support for this theoretical prediction. For example, @sullivanEcosystemStructureEmerges2015 reported FCL is controlled primarily by the number of tributary confluences in Idaho streams. Similarly, it has been shown that the merging of major tributaries lengthened FCL in the Colorado River, especially during a post-monsoon turbid water phase [@saboPulsedFlowsTributary2018]. Clearly, more empirical research is needed to confirm our theoretical predictions.

The interactive effect of disturbance and ecosystem structure has important implications for ecosystem management because anthropogenic climate change is predicted to increase the frequency and magnitude of disturbances [@brunnerFutureStreamflowRegime2020; @mittalImpactHumanIntervention2016]. River networks are especially vulnerable due to their propensity to go dry during droughts [@zipperPervasiveChangesStream2021] as well as susceptibility to more intense flood disturbance with increasing storm intensity [@schiermeierIncreasedFloodRisk2011]. Therefore, the preservation of habitat diversity conferred by ecosystem size and complexity is critical to biodiversity conservation under rapid environmental changes. In addition, maintaining habitat connectivity is particularly important for large, linearly-arranged networks. Although dispersal has received less attention in food chain research [but see @pillaiMetacommunityTheoryExplains2011; @takimotoEffectsProductivityDisturbance2012a; @wangMetapopulationCapacityDetermines2021], our simulation highlighted food webs in these networks are more vulnerable to habitat fragmentation because habitat patches are inherently isolated from one another. Dams and road crossings have altered the connectivity between suitable habitat patches, or have completely blocked dispersal ability [@ishiyamaPredictingEcologicalImpacts2018; @moritaEffectsHabitatFragmentation2002; @perkinFragmentationAltersStream2012a]. Light pollution has also negatively affected the ability of adult aquatic insects to disperse between habitat patches [@perkinEffectsArtificialLighting2014]. These human activities may have disproportionate impacts on food webs in large watersheds with simple network structure.

It is not surprising that FCL declined with increasing omnivory because it effectively compresses a food web (**Figure 1C**) and/or intensifies competition between consumer and predator species. However, how omnivory influences the associations with ecosystem structure has rarely been explored [@takimotoEffectsProductivityDisturbance2012a; @wardMechanisticTheoryAquatic2017]. An important result in our simulation is that omnivory does not change the qualitative relationships between FCL and ecosystem properties, as opposed to productivity effects. This provides a possible explanation for why the ecosystem size hypothesis has been consistently supported in various ecosystems [@takimotoEnvironmentalDeterminantsFoodchain2013] and raises the possibility of ecosystem complexity as an important control of FCL in branching networks. In addition, our results imply that the size and complexity effects are robust to food web reorganization, which is often associated with species extirpations or introductions. The strength of omnivory in these novel communities can be increased or decreased, depending on the introduced species and the extant species pool or community. For example, non-indigenous fishes in the Eastern Mediterranean Sea outcompete native species for high quality prey items, causing native species to increase consumption of non-preferred prey items [@fanelliDepictingNovelEastern2015]. Such substantial changes in food web structure, however, may have limited influences on how ecosystem structure dictates FCL.

Our findings add an important distinction in the discussion of the suitability of assigning protection or conservation status to regions. Current conservation schemes generally emphasize size, aiming to protect the largest areas possible. While this is certainly a useful attribute generally, designating networks with a high degree of variability, or many small patches, can be more beneficial than a large, homogenous area [@fahrigWhySeveralSmall2020]. Accounting for ecosystem complexity is particularly suited to ecosystem architectures in which branches have variable environmental conditions [@koningNetworkGrassrootsReserves2020; @teruiEmergentDualScaling2021]. Conservation and management activities should focus on activities that increase the habitat heterogeneity and connectivity of suitable habitat patches to increase resilience to predicted increases in disturbance frequency and magnitude.

It is abundantly clear that ecosystem complexity in natural systems is the rule, not the exception. Early theoretical and empirical work has assumed simple habitat structure as a necessity. However, habitat heterogeneity, such as the presence of prey refugia, and dispersal between connected habitat patches allow communities to stabilize and persist through space and time [@pedersenNonhierarchicalDispersalPromotes2016; @mougiFoodwebComplexityMetacommunity2016]. Complex ecosystem structure is an extrinsic factor that can affect FCL in communities. Incorporating a realistic understanding of ecosystem structural complexity may improve our perception of communities in naturally complex systems.
